---
title: "American Plaice (\\protect\\textit{Hippoglossoides platessoides}) \\linebreak Snapshot Ecosystem & Socioeconomic Profile"
format: 
  pdf:
    margin-left: "0.5in"
    margin-right: "0.5in"
    margin-top: "0in"
    margin-bottom: "0.5in"
    fontsize: "10pt"
    linestretch: 1
    indent: true
    include-in-header: template.tex    
    # keep-tex: true
    papersize: letter
params:
  widths: !expr c(0.9, 0.75, 3, 3)
  img1: !expr here::here("05_images/noaa_logo.png")
  img2: !expr here::here("05_images/noaa_logo.png")
  img3: !expr here::here("05_images/noaa_logo.png")
  tbl_file: !expr here::here("06_docs/table_template.csv")
  tbl_file2: !expr here::here("06_docs/table_template_2.csv")
  img_dir: !expr here::here("05_images")
  terminal_year: 2026
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
``` 
\backgroundsetup{
  scale=1,
  angle=0,
  opacity=0,
  contents={\includegraphics[width=\paperwidth,height=\paperheight]{bg_pg1.jpg}}
 }
\BgThispage


\vspace{-2.5cm}

\section{Fall `r params$terminal_year`}


:::: {layout="[ [0.4, 0.025, 0.575], [1], [0.575, 0.025, 0.4]] "}

::: {}
<!-- \begin{frame} -->
![](`r params$img1`)
<!-- \end{frame} -->
:::

::: {}
\hfill
:::

::: {}
\section{`r params$terminal_year` in Review}

\subsection{Fishing Community Observations}
- Steady or increasing availability 
- Expanding distributions and changes in migration timing
- Local regulatory complexity affects fishing opportunities

\subsection{Commercial Fishery}
- Number of active vessels declined in 2024, but total landings increased from 2023 
- Total revenue decreased slightly along with average prices ($/lb)
- Average revenue per vessel increased, following an upward trend over the past three years for vessels that remain in the fishery

\subsection{Recreational Fishery}
- Number of targeted trips, catch, and landings all down from 2023
- But number of trips still above the historic average 
- Not clear if catch per angler has continued to increase in 2024

\subsection{Ecosystem}
- Cold winter in the north but near average in the south
- Poor or below average fish condition in recent years
:::

::: {}
\vspace{0.5cm}
\section{Key Points from the Mid-Atlantic Risk Assessment}
:::

::: {}
\vspace{-3cm}

\raggedright
<!-- maybe this is being centered because it's the only section without a header? -->
According to the [Mid-Atlantic `r params$terminal_year` EAFM risk assessment update](`r params$risk_url`), Black Sea Bass scored high and/or moderately high risk in the following elements:

- Moderate-high to high risk to the stock due to:
  - Very high exposure to changes in climate 
  - Observed and potential changes in distribution; northward shift into the Gulf of Maine
  - Dependence on threatened estuarine habitat
  - Decline in the biomass of benthic invertebrate prey 
  - Decline in black sea bass body condition in the Mid Atlantic Bight
- High risk to the recreational fishery due to:
  - Catch exceeding harvest limits in several years
  - High regulatory complexity; frequent changes and varying interstate regulations; regulatory changes in allocations
- Moderate-high risk to the commercial fishery due to:
  - Commercial revenue in wind development areas 
  - High discards & discard mortality

:::

::: {}
\hfill
:::

::: {}
![](`r params$img2`)
![](`r params$img3`)
:::

::::

\newpage

\backgroundsetup{
  contents={\includegraphics[width=\paperwidth,height=\paperheight]{bg_pg2.jpg}}
  }
\BgThispage
<!-- not sure why this doesn't work -->

\newgeometry{top=0.25in, left=0.25in, right=0.25in, bottom=0.25in}

```{r}
data1 <- NEesp2::format_tbl_data(file = params$tbl_file,
                            term_year = params$terminal_year,
                            dir = params$img_dir)
```

::: {}
```{r}
NEesp2::rpt_card_table(data1,
                       widths = params$widths)
```
:::

\newpage

\backgroundsetup{
  contents={\includegraphics[width=\paperwidth,height=\paperheight]{bg_pg2.jpg}}
  }
\BgThispage
<!-- not sure why this doesn't work -->

\newgeometry{top=0.25in, left=0.25in, right=0.25in, bottom=0.25in}

```{r}
data2 <- NEesp2::format_tbl_data(file = params$tbl_file2,
                            term_year = params$terminal_year,
                            dir = params$img_dir)
```

::: {}
```{r}
NEesp2::rpt_card_table(data2,
                       widths = params$widths)
```

\centering
Please contact [nefsc.esp.leads@noaa.gov](nefsc.esp.leads@noaa.gov) with any questions or comments.
:::
