---
title: "American Plaice (\\protect\\textit{Hippoglossoides platessoides}) \\linebreak Snapshot Ecosystem & Socioeconomic Profile"
format: 
  pdf:
    margin-left: "0.5in"
    margin-right: "0.5in"
    margin-top: "0in"
    margin-bottom: "0.5in"
    fontsize: "10pt"
    linestretch: 1
    indent: true
    include-in-header: template.tex    
    # keep-tex: true
    papersize: letter
params:
  widths: !expr c(0.9, 0.75, 3, 3)
  img1: !expr here::here("05_images/plaice_map.png")
  img2: !expr here::here("05_images/condition.png")
  img3: !expr here::here("05_images/noaa_logo.png")
  tbl_file: !expr here::here("06_docs/table_template.csv")
  tbl_file2: !expr here::here("06_docs/table_template_2.csv")
  img_dir: !expr here::here("05_images")
  terminal_year: 2026
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
``` 
\backgroundsetup{
  scale=1,
  angle=0,
  opacity=0,
  contents={\includegraphics[width=\paperwidth,height=\paperheight]{bg_pg1.jpg}}
 }
\BgThispage


\vspace{-2.5cm}

\section{Fall `r params$terminal_year`}


:::: {layout="[ [0.4, 0.025, 0.575], [1], [0.575, 0.025, 0.4]] "}

::: {}
<!-- \begin{frame} -->
![](`r params$img1`)
<!-- \end{frame} -->
:::

::: {}
\hfill
:::

::: {}
\section{`r params$terminal_year` in Review}

\subsection{Fishing Community Observations}
- Text here


\subsection{Commercial Fishery}
- Text here


\subsection{Ecosystem}
- Text here

:::

::: {}
\vspace{0.5cm}
\section{Species Condition/Distribution Maps?/Some other title here}
:::

::: {}
\vspace{0.5cm}

\raggedright
Condition of American plaice across the Northeast Shelf exhibited lows in the 2000s to 2010, indicating thinner, lower-quality fish. Condition values have increased overall since 2010, with most recent years having neutral to good condition, and are experiencing fewer years of poor or below-average condition compared to lows in the 2000s. 
:::

::: {}
\hfill
:::

::: {}
![](`r params$img2`)
:::

::::

\newpage

\backgroundsetup{
  contents={\includegraphics[width=\paperwidth,height=\paperheight]{bg_pg2.jpg}}
  }
\BgThispage
<!-- not sure why this doesn't work -->

\newgeometry{top=0.25in, left=0.25in, right=0.25in, bottom=0.25in}

```{r}
data1 <- NEesp2::format_tbl_data(file = params$tbl_file,
                            term_year = params$terminal_year,
                            dir = params$img_dir)
```

::: {}
```{r}
NEesp2::rpt_card_table(data1,
                       widths = params$widths)
```
:::

\newpage

\backgroundsetup{
  contents={\includegraphics[width=\paperwidth,height=\paperheight]{bg_pg2.jpg}}
  }
\BgThispage
<!-- not sure why this doesn't work -->

\newgeometry{top=0.25in, left=0.25in, right=0.25in, bottom=0.25in}

```{r}
data2 <- NEesp2::format_tbl_data(file = params$tbl_file2,
                            term_year = params$terminal_year,
                            dir = params$img_dir)
```

::: {}
```{r}
NEesp2::rpt_card_table(data2,
                       widths = params$widths)
```

\centering
Please contact [nefsc.esp.leads@noaa.gov](nefsc.esp.leads@noaa.gov) with any questions or comments.
:::
